Robust DPG stability II

### 2012-04-03 docs/NewBCRobustness_paper/paper.tex
fixed typos
git-svn-id: https://svn.ices.utexas.edu/repos/camellia@428 36274a68-8941-4c93-80ef-affaed5139da

The finite element/Galerkin method has been widely utilized in engineering to solve partial differential equations related to simulation of solid mechanics problems. For many problems in engineering, the bilinear forms resulting from these partial differential equations are symmetric and coercive (positive-definite), and it is well known that the finite element method produces near-optimal results for such problems. However, standard Bubnov-Galerkin methods tend to perform poorly for the class of partial-differential equations known as singularly perturbed problems. Singularly perturbed problems are problems where a given parameter may be either very small or very large in the context of physical problems.  An additional complication of singularly perturbed problems is that very often, in the limiting case of the parameter blowing up or decreasing to zero, the partial differential equation itself will change form. A canonical example is the convection-diffusion equation. In 1D, this equation is
\[
u'-\epsilon u'' = f.
\]
The equation represents the change in the concentration of some quantity $u$ in some medium from both convective and diffusive effects. The parameter $\epsilon$ represents the diffusivity, and in the limit of an inviscid medium as $\epsilon\rightarrow 0$, the partial differential equation changes types, going from a parabolic to a hyperbolic equation.

The finite element method has notoriously performed poorly for this example, and for many other singularly perturbed equations. This poor performance is captured by the error bound for the finite element method applied to convection-diffusion equations
\[
\|u-u_h\|_E \leq C(\epsilon) \inf_{w_h}\|u-w_h\|_U,
\]
where $C(\epsilon)$ grows as $\epsilon$ approaches its limiting case. The growth of $C$ with $\epsilon$ is referred to as a \textit{loss of robustness} in $\epsilon$, where the bound on the error in the finite element solution by the best approximation error increases as $\epsilon\rightarrow 0$.  Intuitively, as our singular perturbation parameter changes, our finite element error is bounded more and more loosely by the best approximation error, allowing for more and more error in the finite element solution for as $\epsilon$ shrinks.

A well-known example of this loss of robustness is the growth of spurious oscillations in the finite element solution of the 1D convection-diffusion problem $ u'-\epsilon u'' = f$ (with Dirichlet boundary conditions) as $\epsilon\rightarrow 0$.  For small values of $\epsilon$, the Galerkin approximation of the solution to the convection-diffusion equation develops oscillations of magnitude $O(\epsilon^{-1})$. Another class of examples are wave propagation problems, in which the singular perturbation parameter is the wavenumber, $k\rightarrow \infty$.  The lack of robustness in this case manifests as ``pollution" error, a phenomenon where the finite element solution degrades over many wavelengths for high wavenumbers (commonly manifesting as a phase error between the FE solution and the exact solution).

@@

Under the standard conditions for well-posedness of the continuous variational problem, the discrete DPG method delivers the best approximation error in the ``energy norm" 
\[
\|u\|_E = \sup \frac{b(u,v)}{\|v\|_V}.
\]
The above norm can be determined via $\|u\|_E = \|v_u\|_V$, where $v_u$ solves
\[
(v_u,\delta v)_V = b(u,\delta v), \quad \forall v \in V
\]
As a consequence of this fact, without knowing the exact solution, the actual energy error $\|u-u_h\|_E$ is computable through $\|e_h\|_V$, where
\[
\left(e_h,\delta v\right)_V = b(u-u_h,v) = b(u_h,v)-l(v)
\]
This is simply a consequence of the least-squares nature of DPG; the energy error is simply the measure of the residual in the proper norm.

@@

The DPG (discontinuous Petrov-Galerkin) method is the pairing of the concept of locally computable optimal test functions with the so-called ``ultra-weak formulation". For a given operator equation $Au = f$, the ultra-weak formulation is 
\[
b\left(\left(\widehat{u},u\right),v\right) - l(v) = \langle \widehat{u}, [v] \rangle - (u,A^*v) - (f,v) = 0
\]
where $A^*$ is the formal adjoint resulting from integration by parts, and $[v]$ is the jump of $v$ across element boundaries.  Under the ultra-weak formulation, the regularity requirement on each solution variable is relaxed through integration by parts.  Moreover, to maintain conformity while seeking an $L^2$ setting on interior ``field" variables, boundary terms are identified as additional new unknowns.  The result is a hybridized DG method with locally supported optimal test functions.

@@

The duality between test norm and variational form holds for any particular variational form, though it is often difficult to determine analytically the form of $\|v\|_{V,U}$.  Whether a particular choice of norm on $V$ norm is ``good" or not is problem-dependent; however, for the DPG ultra-weak formulation, there is an important canonical test norm that results from a specific induced norm on $U$.

@@

Our aim now is to translate the above abstract language to a concrete model problem in computational fluid dynamics.  We consider the model convection-diffusion problem on domain $\Omega \in \mathbb{R}^d$ with boundary $\Gamma$
\begin{align}
\div (\beta u) - \epsilon\Delta v &= f  \label{primal}
\end{align}
in first order form
\begin{align*}
\div (\beta u - \sigma) &= f  \\
\frac{1}{\epsilon}\sigma - \grad u &= 0
\end{align*}
with the corresponding variational form (modulo application of boundary conditions)
\begin{align*}
b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) = \left(u,\div \tau - \beta \cdot \grad v\right) + \left(\sigma, \epsilon^{-1} \tau + \grad v\right) - \langle \left[\tau_n\right], \widehat{u} \rangle + \langle \widehat{f}_n, \left[v\right] \rangle 
\end{align*}
where $\left[ \tau_n \right]$ and $\left[ v\right]$ denote the jumps in $\tau_n$ and $v$ across an element edge, defined as
\begin{align*}
\left[ \tau_n \right] &= \tau^+ \cdot n^+ + \tau^-\cdot n^-\\
\left[ v\right] &= v^+{\rm sgn}(n^+) + v^-{\rm sgn}(n^-)
\end{align*}
where $n$ is the outward normal vector to an element edge, and ${\rm sgn}(n)$ is 1 if $n$ aligns with a preset global orientation, and $-1$ otherwise. \textcolor{red}{I need to find more simple way to explain sgn}.

@@

The main goal of this paper is to derive a test norm that \textit{does not induce boundary layers in the optimal test functions}, but still induces an energy norm \textit{that does not lose robustness as $\epsilon \rightarrow 0$}.  
%The first condition is satisfied by first decoupling norms on $\left(\tau,v\right)$ such that
%\[
%\left\|\left(\tau,v\right)\right\|_V = \|\tau\|_{H({\rm div},\Omega_h)} +  \|v\|_{H^1(\Omega_h)}.
%\]
%where $\|\tau\|_{H({\rm div},\Omega_h)}$ and $ \|v\|_{H^1(\Omega_h)}$ are both constructed as the sum of norm and seminorms on each respective space.  Ensuring that each seminorm/normed term is of the same order will allow for smooth solutions (under some assumptions on $\beta$).  
The avoidance of boundary layers in the test norm will be described in more detail in the section detailing numerical experiments.  
\textcolor{red}{Why are we allowed to have $\beta\cdot \grad v$ of order 1 while we have other terms of order $O({\rm Pe})$?  Why does this term not induce boundary layers?}. We focus first on showing the condition on robustness. Such a condition is equivalent to showing the robust boundedness of the energy error from above and below by norms on $U$ 
\[
\left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_{U,1} \lesssim  \left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_E \lesssim \left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_{U,2}
\]
where $\lesssim$ denotes a bound by a constant that will not grow as $\epsilon \rightarrow 0$.  
%$\left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_U \lesssim \left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_E$.  
\textcolor{red}{Explain in intuitive terms what the bounds above and below mean.}
Our goal is to explicitly derive the norms $\|\cdot\|_{U,1}$ and $\|\cdot\|_{U,2}$ that characterize this relation, taking into account the limiting case of $\epsilon \rightarrow 0$. 
Our strategy for doing so is broken down into three parts.

>>>>

The finite element/Galerkin method has been widely utilized in engineering to solve partial differential equations related to simulation of solid mechanics problems. For many problems in engineering, the bilinear forms resulting from these partial differential equations are symmetric and coercive (positive-definite), and it is well known that the finite element method produces near-optimal results for such problems. However, standard Bubnov-Galerkin methods tend to perform poorly for the class of partial-differential equations known as singularly perturbed problems. Singularly perturbed problems are problems where a given parameter may be either very small or very large in the context of physical problems.  An additional complication of singularly perturbed problems is that very often, in the limiting case of the parameter blowing up or decreasing to zero, the partial differential equation itself will change form. A canonical example is the convection-diffusion equation with Dirichlet boundary conditions. In 1D, this equation is
\begin{align*}
u'-\epsilon u'' &= f
\end{align*}
The equation represents the change in the concentration of some quantity $u$ in some medium from both convective and diffusive effects. The parameter $\epsilon$ represents the diffusivity, and in the limit of an inviscid medium as $\epsilon\rightarrow 0$, the partial differential equation changes types, going from a parabolic to a hyperbolic equation. For Dirichlet boundary conditions $u(0)=u_0$ and $u(1)= u_1$, the solution can develop sharp boundary layers of order $\epsilon$ near the outflow.

The finite element method has notoriously performed poorly for both this example, and for many other singularly perturbed problems. This poor performance is captured by the error bound for the finite element method applied to convection-diffusion equations
\[
\|u-u_h\|_E \leq C(\epsilon) \inf_{w_h}\|u-w_h\|_U,
\]
where $C(\epsilon)$ grows as $\epsilon$ approaches its limiting case. The growth of $C$ with $\epsilon$ is referred to as a \textit{loss of robustness} in $\epsilon$, where the bound on the error in the finite element solution by the best approximation error increases as $\epsilon\rightarrow 0$.  Intuitively, as our singular perturbation parameter changes, our finite element error is bounded more and more loosely by the best approximation error, allowing for more and more error in the finite element solution for as $\epsilon$ shrinks. The manifestation of this loss of robustness in the finite element solution of the 1D convection-diffusion problem is the growth of spurious oscillations as $\epsilon\rightarrow 0$.  For a coarse mesh and small values of $\epsilon$, the Galerkin approximation of the solution to the convection-diffusion equation develops oscillations of magnitude $O(\epsilon^{-1})$.

Another example of a singular perturbation problem which experiences loss of robustness is high frequency wave propagation problems, in which the singular perturbation parameter is the wavenumber, $k\rightarrow \infty$. The loss of robustness in this case manifests as ``pollution" error, a phenomenon where the finite element solution degrades over many wavelengths for high wavenumbers (commonly manifesting as a phase error between the FE solution and the exact solution).

@@

Under the standard conditions for well-posedness of the continuous variational problem, the discrete DPG method delivers the best approximation error in the ``energy norm" 
\[
\|u\|_E = \sup \frac{b(u,v)}{\|v\|_V}.
\]
The above norm can be determined via $\|u\|_E = \|v_u\|_V$, where $v_u$ solves
\[
(v_u,\delta v)_V = b(u,\delta v), \quad \forall v \in V
\]
Due to the isometry of the Riesz map, this is equivalent to measuring the norm of the functional $b(u,v) = l_u(v)\in V'$ in the dual norm. An additional consequence of adopting such an energy norm is that, without knowing the exact solution, the actual energy error $\|u-u_h\|_E$ is computable through $\|e_h\|_V$, where
\[
\left(e_h,\delta v\right)_V = b(u-u_h,\delta v) = b(u_h,\delta v)-l(\delta v)
\]
This is simply a consequence of the least-squares nature of DPG; the energy error is simply the measure of the residual in the proper norm - the dual norm on $V$.

@@

The DPG (discontinuous Petrov-Galerkin) method is the pairing of the concept of locally computable optimal test functions with the so-called ``ultra-weak formulation". For a given operator equation in strong form $Au = f$, the ultra-weak formulation is
\[
b\left(\left(\widehat{u},u\right),v\right) - l(v) = \langle \widehat{u}, [v] \rangle - (u,A^*v) - (f,v) = 0
\]
where $A^*$ is the formal adjoint resulting from integration by parts, and $[v]$ is the jump of the test function $v$ across element boundaries.  Both the inner product and formal adjoint are understood in an element-wise fashion. Under the ultra-weak formulation, the regularity requirement on each solution variable is relaxed through integration by parts.  Moreover, to maintain conformity while seeking an $L^2$ setting on interior ``field" variables, boundary terms are identified as additional new unknowns.  The result is a hybridized DG method with locally supported optimal test functions, with energy setting 
\begin{align*}
u&\in L^2(\Omega)\\
v&\in V=D(A^*) \\
\widehat{u}&\in T(D(A))
\end{align*}
where $D(A^*)$ is the domain of the formal adjoint $A^*$, and $T(D(A))$ is the trace space of the domain of the operator $A$ in its strong form. A more comprehensive discussion of the abstract setting of DPG, as well as a proof of well-posedness, is developed using the graph norm in \cite{DPGFriedrichs}.

@@

The duality between test norm and variational form holds for any particular variational form, though it is often difficult to determine analytically the form of $\|v\|_{V,U}$.  Whether a particular choice of norm on $V$ norm is ``good" or not is problem-dependent; however, for the DPG ultra-weak formulation, there is an important canonical test norm that is induced by a specific norm on $U$.

@@

Our aim now is to translate the above abstract language to a concrete model problem in computational fluid dynamics.  We consider the model convection-diffusion problem on domain $\Omega \in \mathbb{R}^d$ with boundary $\Gamma$
\begin{align}
\div (\beta u) - \epsilon\Delta v &= f  \label{primal}
\end{align}
in first order form
\begin{align*}
\div (\beta u - \sigma) &= f  \\
\frac{1}{\epsilon}\sigma - \grad u &= 0
\end{align*}
with the corresponding variational form (prior to the application of boundary conditions)
\begin{align*}
b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) = \left(u,\div \tau - \beta \cdot \grad v\right) + \left(\sigma, \epsilon^{-1} \tau + \grad v\right) - \langle \left[\tau_n\right], \widehat{u} \rangle + \langle \widehat{f}_n, \left[v\right] \rangle 
\end{align*}
where $\left[ \tau_n \right]$ and $\left[ v\right]$ denote the jumps in $\tau_n$ and $v$ across an element edge, defined as
\begin{align*}
\left[ \tau_n \right] &= \tau^+ \cdot n^+ + \tau^-\cdot n^-\\
\left[ v\right] &= v^+{\rm sgn}(n^+) + v^-{\rm sgn}(n^-)
\end{align*}
where $n$ is the outward normal vector to an element edge, and ${\rm sgn}(n)$ is 1 if $n$ aligns with a preset global orientation, and $-1$ otherwise. \textcolor{red}{I need to find more simple way to explain sgn}.

@@

The main goal of this paper is to derive a test norm that \textit{does not induce boundary layers in the optimal test functions}, but still induces an energy norm \textit{that does not lose robustness as $\epsilon \rightarrow 0$}.  
%The first condition is satisfied by first decoupling norms on $\left(\tau,v\right)$ such that
%\[
%\left\|\left(\tau,v\right)\right\|_V = \|\tau\|_{H({\rm div},\Omega_h)} +  \|v\|_{H^1(\Omega_h)}.
%\]
%where $\|\tau\|_{H({\rm div},\Omega_h)}$ and $ \|v\|_{H^1(\Omega_h)}$ are both constructed as the sum of norm and seminorms on each respective space.  Ensuring that each seminorm/normed term is of the same order will allow for smooth solutions (under some assumptions on $\beta$).  
The avoidance of boundary layers in the test norm will be described in more detail in the section detailing numerical experiments.  
\textcolor{red}{Why are we allowed to have $\beta\cdot \grad v$ of order 1 while we have other terms of order $O({\rm Pe})$?  Why does this term not induce boundary layers?}. We focus first on showing the condition on robustness. Such a condition is equivalent to showing the robust boundedness of the energy error from above and below by norms on $U$ 
\[
\left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_{U,1} \lesssim  \left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_E \lesssim \left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_{U,2}
\]
where $\lesssim$ denotes a bound by a constant that will not grow as $\epsilon \rightarrow 0$.  
%$\left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_U \lesssim \left\| \left(u,\sigma,\widehat{u},\widehat{f}_n\right)\right \|_E$.

Intuitively,

\textcolor{red}{Explain in intuitive terms what the bounds above and below mean.}
Our goal is to explicitly derive the norms $\|\cdot\|_{U,1}$ and $\|\cdot\|_{U,2}$ that characterize this relation, taking into account the limiting case of $\epsilon \rightarrow 0$. 
Our strategy for doing so is broken down into three parts.
